\bf Is this normal? A new projection pursuit index to assess a sample against a multivariate null distribution

### 2024-07-04 paper.qmd
small change to title
filled in early example

---
title: "Are you normal? A new projection pursuit index to assess a sample against a multivariate null distribution"
author: "Annalisa Calvi, Ursula Laa, Dianne Cook"
format:
  jasa-pdf:
    keep-tex: true    
bibliography: bibliography.bib  
header-includes: | 
  \usepackage{amsmath}
  \usepackage{amsthm}
  \usepackage{float}
  \usepackage{hyperref}
  \usepackage[utf8]{inputenc}
  \usepackage{bm}
  \def\tightlist{}
  \usepackage{setspace}
  \newcommand\pD{$p\text{-}D$}
  \newcommand\kD{$k\text{-}D$}
  \newcommand\dD{$d\text{-}D$}
  \newcommand\gD{$2\text{-}D$}
  \newtheorem*{theorem}{Theorem}
---

@@

```{r}
#| label: load-libraries
#| warning: false
#| echo: false
library(tourr)
library(mulgar)
library(dplyr)
library(ggplot2)
library(patchwork)
library(knitr)
library(kableExtra)
```

@@

1. Simulating a sample of observations ($\mathbfit{x}$, which are \pD{} vectors)
from $N_p(\mathbfit{\mu}, \Sigma)$. 
2. Transforming each observation to have unit distance from the mean, $\frac{\mathbfit{x}^\top}{||\mathbfit{x}^\top||}$.

To convert this to points on the surface of a confidence ellipsoid,

@@

<PLOTS FROM USCHI's PHYSICS EXAMPLE HERE>

@@

```{r}
#| eval: false
#| echo: false
liver_stats <- tribble(
  ~ALT, ~AST, ~ALP, ~albumin, ~bilirubin, ~GGT, ~protein, ~sex, ~stat,
  5,     5,   30,      35,        0,    2, 60,   "m", "mn",
  40,     30,  120,      50,        20,   44, 80,   "m", "mx",
  5,     5,   30,      35,        0,    2, 60,   "f", "mn",
  40,     30,  120,      50,        20,   70, 80,    "f", "mx"
)
set.seed(1158)
norm_vc <- diag(1, 4, 4)
norm_vc[upper.tri(norm_vc) == TRUE] <- sample(c(0.3, 0.4, 0.5), 6, replace=T)
for (i in 1:3)
  for (j in (i+1):4)
    norm_vc[j, i] <- norm_vc[i, j]
colnames(norm_vc) <- c("GGT", "AST", "ALP", "ALT")
norm_samp <- rmvnorm(500, sigma=as.matrix(norm_vc))
colnames(norm_samp) <- c("GGT", "AST", "ALP", "ALT")
set.seed(217)

women_vc <- norm_vc
women_vc[1,4] <- 0.5
women_vc[4,1] <- 0.5
women <- rmvnorm(16, mean=c(0.3, -0.3, -1.2, -1.5),
                 sigma = women_vc)
colnames(women) <- c("GGT", "AST", "ALP", "ALT")
women <- as.data.frame(women)

animate_xy(women,            
           ellipse=norm_vc, 
           axes = "bottomleft", 
           half_range=5, 
           center=FALSE)

>>>>

---
title: "Is this normal? A new projection pursuit index to assess a sample against a multivariate null distribution"
author: "Annalisa Calvi, Ursula Laa, Dianne Cook"
format:
  jasa-pdf:
    keep-tex: true    
bibliography: bibliography.bib  
header-includes: | 
  \usepackage{amsmath}
  \usepackage{amsthm}
  \usepackage{float}
  \usepackage{hyperref}
  \usepackage[utf8]{inputenc}
  \usepackage{bm}
  \def\tightlist{}
  \usepackage{setspace}
  \newcommand\pD{$p\text{-}D$}
  \newcommand\kD{$k\text{-}D$}
  \newcommand\dD{$d\text{-}D$}
  \newcommand\gD{$2\text{-}D$}
  \newtheorem*{theorem}{Theorem}
---

@@

```{r}
#| label: load-libraries
#| warning: false
#| echo: false
library(tourr)
library(mulgar)
library(dplyr)
library(ggplot2)
library(patchwork)
library(knitr)
library(kableExtra)
library(mvtnorm)
```

@@

1. Simulating a sample of observations ($\mathbfit{x}$, which are \pD{} vectors)
from $N_p(\mathbfit{\mu}, \Sigma)$. 
2. Transforming each observation to have unit distance from the mean, $\frac{\mathbfit{x}^\top}{||\mathbfit{x}^\top||}$. 
To convert this to points on the surface of a confidence ellipsoid,

@@

For example, @fit-compare compares a new sample of patient scores against the normal range represented by an ellipse.

```{r}
#| eval: false
#| echo: false
liver_stats <- tribble(
  ~ALT, ~AST, ~ALP, ~albumin, ~bilirubin, ~GGT, ~protein, ~sex, ~stat,
  5,     5,   30,      35,        0,    2, 60,   "m", "mn",
  40,     30,  120,      50,        20,   44, 80,   "m", "mx",
  5,     5,   30,      35,        0,    2, 60,   "f", "mn",
  40,     30,  120,      50,        20,   70, 80,    "f", "mx"
)
set.seed(1158)
norm_vc <- diag(1, 4, 4)
norm_vc[upper.tri(norm_vc) == TRUE] <- sample(c(0.3, 0.4, 0.5), 6, replace=T)
for (i in 1:3)
  for (j in (i+1):4)
    norm_vc[j, i] <- norm_vc[i, j]
colnames(norm_vc) <- c("GGT", "AST", "ALP", "ALT")
norm_samp <- rmvnorm(500, sigma=as.matrix(norm_vc))
colnames(norm_samp) <- c("GGT", "AST", "ALP", "ALT")
norm_samp <- as.data.frame(norm_samp)
norm_ci <- gen_vc_ellipse(norm_vc) * 2
colnames(norm_ci) <- c("GGT", "AST", "ALP", "ALT")
norm_ci <- as.data.frame(norm_ci)

set.seed(217)
women_vc <- norm_vc
women_vc[1,4] <- 0.5
women_vc[4,1] <- 0.5
women <- rmvnorm(16, mean=c(0.3, -0.3, -1.2, -1.5),
                 sigma = women_vc)
colnames(women) <- c("GGT", "AST", "ALP", "ALT")
women <- as.data.frame(women)

norm_plus <- bind_rows(norm_samp, women)
norm_plus <- norm_plus |>
  mutate(type = factor(c(rep("norm", 500), rep("samp", 16))))
animate_xy(norm_plus[,1:4], col=norm_plus$type)

norm_ci_plus <- bind_rows(norm_ci, women)
norm_ci_plus <- norm_ci_plus |>
  mutate(type = factor(c(rep("ci", 500), rep("samp", 16))))
animate_xy(norm_ci_plus[,1:4], col=norm_ci_plus$type)

::: {#fig-compare layout-ncol=2}

![Sample of patients relative to confidence ellipse.](images/example1.png){#fig-compare1 width=300 align="center"}

![Sample of patients relative to relative to simulated sample of normal patients.](images/example2.png){#fig-compare width=300 align="center"}

Illustration of current procedure: (a) compare new sample with points of the surface of a $p-D$ ellipse, (b) compare new sample with a simulated sample of normal patients. Although both approaches are useful, the rough edges of the projected ellipse points makes it difficult to precisely assess the positions of the new sample against the normal bounds.
:::

@@

```{r}
#| eval: false
#| echo: false
animate_xy(women,            
           ellipse=norm_vc, 
           axes = "bottomleft", 
           half_range=5, 
           center=FALSE)

### 2024-07-04 paper.tex
small change to title
filled in early example

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Are you normal? A new projection pursuit index to assess a sample against a multivariate null distribution},
  pdfauthor={Annalisa Calvi, Ursula Laa, Dianne Cook},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\title{\bf Are you normal? A new projection pursuit index to assess a
sample against a multivariate null distribution}
\author{
Annalisa Calvi, Ursula Laa, Dianne Cook\\
}
\maketitle

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Simulating a sample of observations (\(\mathbfit{x}\), which are \pD{}
  vectors) from \(N_p(\mathbfit{\mu}, \Sigma)\).
\item
  Transforming each observation to have unit distance from the mean,
  \(\frac{\mathbfit{x}^\top}{||\mathbfit{x}^\top||}\).
\end{enumerate}

To convert this to points on the surface of a confidence ellipsoid,

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{2}
\tightlist
\item
  transform the shape using a specific variance covariance, and shift to
  center on the mean vector.
\end{enumerate}

@@

\textless PLOTS FROM USCHI's PHYSICS EXAMPLE HERE\textgreater{}

>>>>

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Is this normal? A new projection pursuit index to assess a sample against a multivariate null distribution},
  pdfauthor={Annalisa Calvi, Ursula Laa, Dianne Cook},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\title{\bf Is this normal? A new projection pursuit index to assess a
sample against a multivariate null distribution}
\author{
Annalisa Calvi, Ursula Laa, Dianne Cook\\
}
\maketitle

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Simulating a sample of observations (\(\mathbfit{x}\), which are \pD{}
  vectors) from \(N_p(\mathbfit{\mu}, \Sigma)\).
\item
  Transforming each observation to have unit distance from the mean,
  \(\frac{\mathbfit{x}^\top}{||\mathbfit{x}^\top||}\). To convert this
  to points on the surface of a confidence ellipsoid,
\item
  transform the shape using a specific variance covariance, and shift to
  center on the mean vector.
\end{enumerate}

@@

For example, \citet{fit-compare} compares a new sample of patient scores
against the normal range represented by an ellipse.

\includegraphics[width=3.125in,height=\textheight]{images/example1.png}

\subcaption{\label{fig-compare1}Sample of patients relative to
confidence ellipse.}

\includegraphics[width=3.125in,height=\textheight]{images/example2.png}

\subcaption{\label{fig-compare}Sample of patients relative to relative
to simulated sample of normal patients.}

\caption{\label{fig-compare}Illustration of current procedure: (a)
compare new sample with points of the surface of a \(p-D\) ellipse, (b)
compare new sample with a simulated sample of normal patients. Although
both approaches are useful, the rough edges of the projected ellipse
points makes it difficult to precisely assess the positions of the new
sample against the normal bounds.}
